---
title: "Bayesian Quadratic-Plateau Modeling for Critical Soil Test Value Estimation"
subtitle: "Posterior Uncertainty and Probability of Response"
author: "Gustavo A. Roa"
date: "2026-03-27"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    self-contained: true
execute:
  warning: false
  message: false
---

# Overview
## Purpose

This tutorial demonstrates how to:

\- Fit a Bayesian quadratic-plateau (B-QP) model.

\- Estimate the critical soil test value (CSTV).

\- Quantify posterior uncertainty.

\- Compute the probability that expected yield gain exceeds a specified threshold.

\- Construct probability tables and response-risk visualizations.

The emphasis is on shifting inference from a single breakpoint estimate to probability-based decision support.

------------------------------------------------------------------------

# Setup
Loads required packages for Bayesian modeling, marginal likelihood estimation, and data manipulation.
```{r}
# Core packages
library(tidyverse)

# Data I/O
library(readxl)

# Bayesian fitting
library(brms)

```


# Data
## Import data

```{r}
dat <- read_excel("./Data_example.xlsx", sheet = "KS_30")

```

## Visual inspection
Before fitting any model, inspect the observed response pattern.
```{r}
ggplot(dat, aes(input_level, response)) +
  geom_point()

```

# Bayesian Quadratic-Plateau Model
## Model specification and fitting

```{r}
# brms formula
bf_QP <- brms::bf(response ~ b0 + b1 * fmin(input_level, xs) - (b1 / (2 * xs)) * 
                    fmin(input_level, xs)^2, b0 + b1 + xs ~ 1, nl = TRUE)

# Priors (paper-centered)
               # Intercept (B0)
priors_QP <- c(brms::set_prior("normal(58.3, 0.5*58.3)", nlpar = "b0", lb=0),
               # Linear term (B1)
               brms::set_prior("normal(3.8, 0.5*3.8)",  nlpar = "b1"),
               # Critical value (xs = theta)
               brms::set_prior("normal(21.4, 0.5*21.4)", nlpar = "xs", lb=0),
               # Residual SD
               brms::set_prior("student_t(3, 0, 10)", class = "sigma"))

# Fit brms model
fit_QP_bayes <- brms::brm(bf_QP, data = dat, prior = priors_QP, chains = 5, 
                          iter = 3000, warmup = 1000, cores = 1, seed = 555,
                          control = list(adapt_delta = 0.95))

fit_QP_bayes


```

# Posterior-Based Yield Gain Inference
Rather than focusing only on the CSTV, we compute the posterior probability that expected yield gain exceeds a threshold.

## Extract Posterior Draws and Define the Quadratic-Plateau Response Model
```{r}
# Extract posterior draws
dr <- as_draws_df(fit_QP_bayes)

# Model parameters
b0  <- dr$b_b0_Intercept
b1  <- dr$b_b1_Intercept
xs <- dr$b_xs_Intercept

# Define QP mean response function
# μ(x)=a+bm−2xsb*m2, m=min(x,xs)
# QP mean function
mu_qp <- function(x, b0, b1, xs){
  m <- pmin(x, xs)
  b0 + b1*m - (b1/(2*xs))*m^2
}

# Plateau mean (x = xs)
mu_plateau <- function(b0, b1, xs){
  b0 + (b1*xs)/2
}

# Posterior draws of yield gain
# Δ(x)=μ(xs)−μ(x)
gain_draws_mean <- function(x0){
  mu_plateau(b0, b1, xs) - mu_qp(x0, b0, b1, xs)
}

# Probability of gain exceeding threshold δ
p_gain_mean <- function(x0, delta = 0){
  g <- gain_draws_mean(x0)
  mean(g > delta, na.rm = TRUE)
}
```

## Compute the Probability of Response Across Soil Test Levels
```{r}
# Compute response probability curve
x_grid <- seq(1, 50, by = 0.1)

# Probability expected gain > 0
p0 <- sapply(x_grid, p_gain_mean, delta = 0) # p0 now contains: P(Δ(x)>0)

#Define risk thresholds
p_thr <- c(0.70, 0.30)

# Find soil test boundaries where probabilities cross thresholds
find_boundary <- function(x, p, thr){
  idx <- which(p <= thr)[1]
  if (is.na(idx)) return(Inf)
  if (idx == 1) return(x[1])

  x1 <- x[idx-1]; x2 <- x[idx]
  p1 <- p[idx-1]; p2 <- p[idx]

  x1 + (thr - p1) * (x2 - x1) / (p2 - p1)
}

bounds <- sapply(p_thr, function(thr) find_boundary(x_grid, p0, thr))

# Build table
risk <- data.frame(
  Class = c("High response probability",
            "Moderate response probability",
            "Low response probability"),
  Prob_range = c("[0.70, 1.00]",
                 "[0.30, 0.70)",
                 "[0.00, 0.30)"),
  lower = c(-Inf, bounds), upper = c(bounds, Inf))

risk 

```

## Estimate Probability of Response Within Mehlich-3 Soil Test Categories
```{r}
# Build dataframe with soil test and probability
prob_df <- data.frame(
  Soil_Test_ppm = x_grid,
  Prob_response = p0
)

# Mehlich-3 categories
cat_table <- data.frame(
  Category = c("Very Low","Low","Medium","High","Very High"),
  lower = c(0, 9, 16, 21, 31),
  upper = c(8, 15, 20, 30, Inf)
)

# Compute mean probability of response within each category
cat_table$Prob_response <- sapply(1:nrow(cat_table), function(i){
  idx <- prob_df$Soil_Test_ppm >= cat_table$lower[i] &
         prob_df$Soil_Test_ppm <= cat_table$upper[i]
  mean(prob_df$Prob_response[idx], na.rm = TRUE)
})

cat_table

```

```{r}
cat_table$Prob_min <- sapply(1:nrow(cat_table), function(i){
  idx <- prob_df$Soil_Test_ppm >= cat_table$lower[i] &
         prob_df$Soil_Test_ppm <= cat_table$upper[i]
  min(prob_df$Prob_response[idx], na.rm = TRUE)
})

cat_table$Prob_max <- sapply(1:nrow(cat_table), function(i){
  idx <- prob_df$Soil_Test_ppm >= cat_table$lower[i] &
         prob_df$Soil_Test_ppm <= cat_table$upper[i]
  max(prob_df$Prob_response[idx], na.rm = TRUE)
})

cat_table

```


# Probability Lookup Table
```{r}
# Define the table grid you want
x_vals     <- seq(0, 25, by = 1)     # soil test ppm rows: 2,4,...,28
delta_vals <- seq(0, 10, by = 1)     # delta columns: 2,4,...,20 (RY points)

# Compute
grid <- expand.grid(Soil_Test_ppm = x_vals, Delta_RY = delta_vals)

grid$Prob_response <- mapply(FUN = p_gain_mean, x0 = grid$Soil_Test_ppm, 
                             delta = grid$Delta_RY)

```

# Probability Surface

```{r fig.height=6, fig.width=10, warning=FALSE}
heat_df <- grid %>%
  mutate(Prob_pct = 100 * Prob_response)

ggplot(heat_df,
                  aes(x = Soil_Test_ppm,
                      y = Delta_RY,
                      z = Prob_pct)) +
  
  # Filled contour surface
  geom_contour_filled(breaks = seq(0, 100, 10)) +
  
  # Scales and theme
  scale_x_continuous(limits = c(0, 25), breaks = seq(0, 25, 5), expand = c(0,0)) +
  scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 2), expand = c(0,0)) +
  
  scale_fill_viridis_d(name = "Probability (%)", 
                       labels = paste(seq(0,90,10), seq(10,100,10), sep = "–"),
                       option = "D", alpha=0.8) +
  
  labs(y = "ΔRY", x = expression(Soil~test~P~(mg~kg^-1))) +

  theme_bw(base_size = 22) +
  theme(panel.grid = element_blank(), axis.text = element_text(color = "black"),
        panel.background = element_rect(color = "white", linewidth = 1.5)) 

```


# Probability Curves for Selected Gain Thresholds
```{r fig.height=6, fig.width=10, warning=FALSE}
delta_selected <- c(1, 3, 5)

curve_df <- expand.grid(Soil_Test_ppm = x_grid, Delta_RY = delta_selected)

curve_df$Prob <- mapply(p_gain_mean, x0 = curve_df$Soil_Test_ppm,
                        delta = curve_df$Delta_RY)

ggplot(curve_df, aes(x = Soil_Test_ppm, y = 100*Prob, color=factor(Delta_RY))) +
  geom_line(linewidth = 1) + 
  scale_color_viridis_d(name = "ΔRY ≥", option = "E") +
  
  scale_x_continuous(limits = c(0, 30), breaks = seq(0, 30, 5)) +
  
  labs(x = expression(Soil~test~P~(mg~kg^-1)), 
       y = "Posterior probability of gain (%)") +
  theme_bw(base_size = 22) +
  theme(panel.grid = element_blank(), axis.text = element_text(color = "black"),
        panel.background = element_rect(color = "white", linewidth = 1.5)) 

```



